---
title: "Building a Global Datacenter Dataset: Malaysia Pilot"
author: "Daniel"
date: today
format:
  pdf:
    toc: true
    number-sections: true
    geometry: margin=1in
    fig-width: 7
    fig-height: 5
engine: knitr
execute:
  warning: false
  message: false
python:
  executable: /opt/anaconda3/envs/textmining/bin/python
---

# Project Overview

I wanted to build a replicable pipeline for constructing a global datacenter
dataset from openly-licensed sources. Commercial directories like Data Center
Map prohibit scraping and manual extraction in their terms of service, so I
built this around three open sources instead: OpenStreetMap via the Overpass
API, Wikidata via SPARQL, and cloud provider region endpoints. This document
implements the pipeline for a single country — Malaysia — as a pilot. Malaysia
was chosen deliberately: Johor has become one of the most active hyperscale
build-out markets in Southeast Asia since 2023, which means the country is
small enough to run end-to-end in seconds but large enough to produce an
interesting dataset with real analytical value.

The deliverable is a tidy dataset of datacenter facilities in Malaysia with
names, operators where available, coordinates, source attribution, and
country-level metadata joined from Natural Earth shapefiles. The pipeline mixes
R and Python via reticulate, following the same pattern I used in Labs 7 and 8.

# Part 1: Environment Setup

## Configure reticulate and load R packages

I began by pointing reticulate at the `textmining` conda environment before
loading any packages that touch Python. Setting `RETICULATE_PYTHON` as an
environment variable belt-and-suspenders alongside the YAML `python:` field is
the pattern that worked reliably in Lab 8, especially when spacyr is in the
mix. I then loaded the R packages needed for HTTP, JSON parsing, spatial
operations, and mapping.

```{r}
#| label: setup-r
#| message: false
#| warning: false

Sys.setenv(RETICULATE_PYTHON = "/opt/anaconda3/envs/textmining/bin/python")

library(reticulate)
use_condaenv("textmining", required = TRUE)

library(httr)
library(jsonlite)
library(dplyr)
library(tidyr)
library(stringr)
library(purrr)
library(sf)
library(leaflet)
library(ggplot2)
library(knitr)
```

## Verify Python environment

I confirmed reticulate was using the correct Python binary before moving on.
This is the check that caught environment drift problems in Lab 10.

```{r}
#| label: verify-python
py_config()
```

# Part 2: Query OpenStreetMap via Overpass

## Build the Overpass query

OpenStreetMap contributors tag datacenters using two main schemas:
`telecom=data_center` (the canonical tag) and `building=data_center` (used
when the tag is attached to the building geometry rather than an amenity
node). I queried for both, scoped to Malaysia using the ISO country code
`MY`. The `[out:json]` header returns JSON, `out center tags` asks Overpass
to return the centroid coordinates for ways and relations rather than their
full geometry, which keeps the payload small.

```{r}
#| label: overpass-query

overpass_query <- paste(
  '[out:json][timeout:180];',
  'area["ISO3166-1"="MY"]->.my;',
  '(',
  '  node["telecom"="data_center"](area.my);',
  '  way["telecom"="data_center"](area.my);',
  '  relation["telecom"="data_center"](area.my);',
  '  node["building"="data_center"](area.my);',
  '  way["building"="data_center"](area.my);',
  '  relation["building"="data_center"](area.my);',
  ');',
  'out center tags;',
  sep = "\n"
)

cat(overpass_query)
```

## Send the request

I sent the query to the main Overpass endpoint. The response comes back as
JSON with an `elements` array where each element is a node, way, or relation
with its tags and coordinates. I added a user-agent header as a courtesy —
Overpass operators appreciate knowing who is hitting their endpoint, and it
reduces the chance of being throttled.

```{r}
#| label: overpass-request

overpass_resp <- POST(
  "https://overpass-api.de/api/interpreter",
  body = list(data = overpass_query),
  encode = "form",
  add_headers(`User-Agent` = "ITEC724-DatacenterPilot/1.0"),
  timeout(200)
)

cat("Status:", status_code(overpass_resp), "\n")

raw <- content(overpass_resp, as = "text", encoding = "UTF-8")

if (startsWith(trimws(raw), "<")) {
  cat("Overpass returned an error page. First 1000 chars:\n")
  cat(substr(raw, 1, 1000))
  stop("Overpass returned HTML/XML instead of JSON")
}

if (!dir.exists("outputs")) dir.create("outputs")
writeLines(raw, "outputs/overpass_malaysia_raw.json")
cat("Saved raw response to outputs/overpass_malaysia_raw.json\n")
cat("First 300 chars:\n")
cat(substr(raw, 1, 300))
```

## Parse the response into a tibble

The raw JSON needed flattening. Nodes have `lat`/`lon` directly; ways and
relations have their centroid nested under `center`. I unified both into a
single tibble with consistent columns and pulled the most useful tags out of
the nested `tags` object — name, operator, and the two datacenter-identifying
tags. Any row missing both a name and an operator got a synthetic label so
downstream joins still work.

```{r}
#| label: parse-overpass

library(jsonlite)
library(dplyr)
library(knitr)

if (!exists("raw") || !is.character(raw)) {
  raw <- readLines("outputs/overpass_malaysia_raw.json") |> paste(collapse = "\n")
}

parsed <- fromJSON(raw, flatten = TRUE)

elements <- parsed$elements

get_col <- function(df, colname) {
  if (colname %in% names(df)) df[[colname]] else rep(NA, nrow(df))
}

osm_df <- tibble(
  osm_id   = elements$id,
  osm_type = elements$type,
  lat      = ifelse(is.na(get_col(elements, "lat")),
                    get_col(elements, "center.lat"),
                    get_col(elements, "lat")),
  lon      = ifelse(is.na(get_col(elements, "lon")),
                    get_col(elements, "center.lon"),
                    get_col(elements, "lon")),
  name     = get_col(elements, "tags.name"),
  operator = get_col(elements, "tags.operator"),
  telecom  = get_col(elements, "tags.telecom"),
  building = get_col(elements, "tags.building")
) %>%
  mutate(
    name = ifelse(is.na(name) | name == "",
                  paste0("Unnamed facility (OSM ", osm_type, " ", osm_id, ")"),
                  name),
    source = "OpenStreetMap"
  ) %>%
  filter(!is.na(lat), !is.na(lon))

cat("Parsed", nrow(osm_df), "datacenter facilities from OSM\n\n")
kable(head(osm_df, 10), caption = "First 10 datacenters from OpenStreetMap")
```

# Part 3: Query Wikidata via SPARQL

## Build the SPARQL query

Wikidata's structured data sits behind a SPARQL endpoint. I queried for any
item that is an `instance of` data center (Q1149652) or a subclass thereof,
located in Malaysia (Q833), and with geographic coordinates. The `SERVICE
wikibase:label` block asks Wikidata to return human-readable labels in
English, which saves a separate lookup for each item.

```{r}
#| label: wikidata-query

sparql_query <- '
SELECT ?item ?itemLabel ?coords ?operatorLabel ?inceptionDate WHERE {
  ?item wdt:P31/wdt:P279* wd:Q1149652 .
  ?item wdt:P17 wd:Q833 .
  ?item wdt:P625 ?coords .
  OPTIONAL { ?item wdt:P137 ?operator . }
  OPTIONAL { ?item wdt:P571 ?inceptionDate . }
  SERVICE wikibase:label { bd:serviceParam wikibase:language "en" . }
}
'

cat(sparql_query)
```

## Send the SPARQL request

Wikidata's query service accepts GET requests with the query as a URL
parameter and returns JSON when the `Accept` header is set appropriately.

```{r}
#| label: wikidata-request

wd_response <- GET(
  url = "https://query.wikidata.org/sparql",
  query = list(query = sparql_query, format = "json"),
  add_headers(
    `Accept`     = "application/sparql-results+json",
    `User-Agent` = "ITEC724-DatacenterPilot/1.0 (academic research)"
  )
)

status_code(wd_response)
```

## Parse the SPARQL response

Wikidata returns coordinates as WKT-formatted strings like
`Point(lon lat)`, which I parsed out with a regex. The result is a tidy
tibble matching the OSM schema so the two sources can be unioned.

```{r}
#| label: parse-wikidata

wd_raw <- content(wd_response, as = "text", encoding = "UTF-8")
wd_parsed <- fromJSON(wd_raw, flatten = TRUE)

bindings <- wd_parsed$results$bindings

safe_col <- function(df, colname) {
  if (is.data.frame(df) && colname %in% names(df)) df[[colname]] else NA_character_
}

if (!is.null(bindings) && length(bindings) > 0 && nrow(bindings) > 0) {
  
  wd_df <- tibble(
    wikidata_id = basename(safe_col(bindings, "item.value")),
    name        = safe_col(bindings, "itemLabel.value"),
    coords_wkt  = safe_col(bindings, "coords.value"),
    operator    = safe_col(bindings, "operatorLabel.value"),
    inception   = safe_col(bindings, "inceptionDate.value")
  ) %>%
    mutate(
      lon = as.numeric(str_extract(coords_wkt, "(?<=Point\\()-?[0-9\\.]+")),
      lat = as.numeric(str_extract(coords_wkt, "-?[0-9\\.]+(?=\\))")),
      source = "Wikidata"
    ) %>%
    select(-coords_wkt)
  
  cat("Retrieved", nrow(wd_df), "datacenters from Wikidata\n")
  
} else {
  
  wd_df <- tibble(
    wikidata_id = character(),
    name = character(),
    operator = character(),
    inception = character(),
    lon = numeric(),
    lat = numeric(),
    source = character()
  )
  
  cat("Wikidata returned no results for Malaysia\n")
}

if (nrow(wd_df) > 0) {
  kable(wd_df, caption = "Datacenters from Wikidata")
} else {
  cat("(Empty table — dedup step will skip Wikidata source)\n")
}
```



```{r}
#| label: stage-shapefiles

if (!dir.exists("data")) dir.create("data")
if (!dir.exists("data/world-shape-files")) dir.create("data/world-shape-files")

src_dir <- "lab-8-collaborative-data-visualization-psg2/data/world-shape-files"
dst_dir <- "data/world-shape-files"

files_to_copy <- list.files(src_dir, pattern = "^ne_50m_admin_0_countries",
                            full.names = TRUE)

file.copy(files_to_copy, dst_dir, overwrite = TRUE)

list.files(dst_dir, pattern = "^ne_50m_admin_0_countries")
```

# Part 4: Merge and Deduplicate

## Union the two sources

I combined the OSM and Wikidata tibbles, keeping only the columns common to
both plus source-specific identifiers. Since Malaysia is small and the two
sources have different biases — OSM is broader and community-mapped, Wikidata
is narrower and editor-curated — I expected some overlap on the largest named
facilities.

```{r}
#| label: union-sources

osm_slim <- osm_df %>%
  select(name, operator, lat, lon, source, osm_id, osm_type)

wd_slim <- wd_df %>%
  select(name, operator, lat, lon, source, wikidata_id)

combined <- bind_rows(osm_slim, wd_slim) %>%
  mutate(
    osm_id       = as.character(osm_id),
    wikidata_id  = as.character(wikidata_id)
  )

cat("Total rows before dedup:", nrow(combined), "\n")
```

## Deduplicate by spatial proximity

A facility appearing in both sources will have very similar coordinates but
may have slightly different names (OSM often has the building name, Wikidata
the corporate name). I deduplicated by rounding coordinates to three decimal
places — roughly 111 meters at the equator — and collapsing rows that match
on rounded location. When a row was present in both sources I kept the
Wikidata name (usually more canonical) and merged the identifiers.

```{r}
#| label: dedup

deduped <- combined %>%
  mutate(
    lat_key = round(lat, 3),
    lon_key = round(lon, 3)
  ) %>%
  group_by(lat_key, lon_key) %>%
  summarise(
    name = {
      wd_names <- name[source == "Wikidata"]
      osm_names <- name[source == "OpenStreetMap"]
      if (length(wd_names) > 0 && !is.na(wd_names[1])) wd_names[1]
      else if (length(osm_names) > 0) osm_names[1]
      else NA_character_
    },
    operator = {
      non_na_ops <- operator[!is.na(operator)]
      if (length(non_na_ops) > 0) non_na_ops[1] else NA_character_
    },
    lat = mean(lat, na.rm = TRUE),
    lon = mean(lon, na.rm = TRUE),
    sources = paste(unique(source), collapse = "; "),
    osm_id = paste(na.omit(unique(osm_id)), collapse = "; "),
    wikidata_id = paste(na.omit(unique(wikidata_id)), collapse = "; "),
    n_sources = n_distinct(source),
    .groups = "drop"
  ) %>%
  select(-lat_key, -lon_key) %>%
  arrange(desc(n_sources), name)

cat("Total facilities after dedup:", nrow(deduped), "\n")
cat("Facilities in multiple sources:", sum(deduped$n_sources > 1), "\n")
```

# Part 5: Spatial Enrichment with Natural Earth

## Load country boundaries and join

I used the Natural Earth 50m country shapefile that was already staged in
the project from `get-shapefiles.R`. Converting the facility tibble to an
`sf` object with CRS 4326 (WGS84) let me spatial-join against the country
polygon to confirm Malaysia membership and pull ISO codes and region
classifications into the dataset.

```{r}
#| label: spatial-join

library(sf)
library(dplyr)
library(knitr)

ne_path <- "data/world-shape-files/ne_50m_admin_0_countries.shp"

if (file.exists(ne_path)) {
  countries <- st_read(ne_path, quiet = TRUE)

  facilities_sf <- deduped %>%
    filter(!is.na(lat), !is.na(lon)) %>%
    st_as_sf(coords = c("lon", "lat"), crs = 4326, remove = FALSE)

  enriched <- st_join(
    facilities_sf,
    countries %>% select(ADMIN, ISO_A3, CONTINENT, SUBREGION),
    join = st_intersects
  ) %>%
    st_drop_geometry() %>%
    as_tibble()
} else {
  warning("Natural Earth shapefile not found; skipping spatial join")
  enriched <- deduped %>%
    mutate(ADMIN = "Malaysia", ISO_A3 = "MYS",
           CONTINENT = "Asia", SUBREGION = "South-Eastern Asia")
}

kable(head(enriched, 10),
      caption = "First 10 enriched facilities with country metadata")
```

# Part 6: Python Enrichment via reticulate

## Switch to Python for text normalization

I wanted to demonstrate the R-Python bridge pattern from Lab 10 by doing name
normalization in Python using pandas. This is admittedly overkill for a
single-country pilot — `stringr` would handle it fine in R — but the pattern
matters for when the pipeline scales globally and I want to plug in spaCy
entity extraction on facility descriptions.

```{python}
#| label: python-normalize

import pandas as pd
import re

df = r.enriched.copy()

def normalize_name(s):
    if pd.isna(s):
        return s
    s = str(s).strip()
    s = re.sub(r'\s+', ' ', s)
    s = re.sub(r'\bData\s*Cent(er|re)\b', 'Data Center', s, flags=re.IGNORECASE)
    s = re.sub(r'\bDC\b', 'Data Center', s)
    return s

df['name_normalized'] = df['name'].apply(normalize_name)

df['has_operator']   = df['operator'].notna()
df['in_both_sources'] = df['n_sources'] > 1

print(f"Total facilities: {len(df)}")
print(f"With known operator: {df['has_operator'].sum()}")
print(f"Appearing in multiple sources: {df['in_both_sources'].sum()}")

enriched_py = df
```

## Pull the Python dataframe back into R

I passed the normalized dataframe back to R using `py$` access so downstream
mapping and export stay in the R side of the pipeline.

```{r}
#| label: python-back-to-r

final_df <- py$enriched_py %>% as_tibble()

kable(
  final_df %>% select(name_normalized, operator, lat, lon, sources, ISO_A3),
  caption = "Final Malaysia datacenter dataset"
)
```

# Part 7: Visualization

## Interactive leaflet map

Following the Lab 8 pattern, I built a leaflet map with clustered markers
and popups showing facility metadata. This renders interactively in HTML
and falls back to a webshot PNG when rendering to PDF.

```{r}
#| label: leaflet-map

leaflet_map <- leaflet(final_df) %>%
  addTiles() %>%
  addCircleMarkers(
    lng = ~lon, lat = ~lat,
    radius = 6,
    color = "red", fillOpacity = 0.7,
    popup = ~paste0(
      "<strong>", name_normalized, "</strong><br>",
      "Operator: ", ifelse(is.na(operator), "Unknown", operator), "<br>",
      "Sources: ", sources
    )
  ) %>%
  setView(lng = 109, lat = 4, zoom = 6)

if (knitr::is_latex_output()) {
  if (!dir.exists("figures")) dir.create("figures")
  html_file <- file.path("figures", "malaysia_dc_map.html")
  png_file  <- file.path("figures", "malaysia_dc_map.png")
  htmlwidgets::saveWidget(leaflet_map, file = html_file, selfcontained = TRUE)
  webshot::webshot(html_file, file = png_file, vwidth = 900, vheight = 600)
  knitr::include_graphics(png_file)
} else {
  leaflet_map
}
```

## Source coverage summary

I wrapped up with a quick ggplot summary showing how many facilities came
from each source and how many appeared in both. This is the kind of
provenance transparency that matters when the dataset eventually gets
cited.

```{r}
#| label: coverage-plot
#| fig-width: 6
#| fig-height: 4

coverage <- final_df %>%
  count(sources, name = "n_facilities") %>%
  arrange(desc(n_facilities))

ggplot(coverage, aes(x = reorder(sources, n_facilities), y = n_facilities)) +
  geom_col(fill = "steelblue") +
  coord_flip() +
  labs(
    title    = "Datacenter facility coverage by source",
    subtitle = "Malaysia pilot",
    x        = "Source(s)",
    y        = "Number of facilities"
  ) +
  theme_minimal()
```

# Part 8: Export

## Write the final dataset

I saved the final tibble as both CSV (for portability) and RDS (for
round-tripping without losing types). The `outputs/` folder is where every
artifact from this pipeline gets staged.

```{r}
#| label: export

if (!dir.exists("outputs")) dir.create("outputs")

write.csv(final_df,
          file = "outputs/malaysia_datacenters.csv",
          row.names = FALSE)
saveRDS(final_df, file = "outputs/malaysia_datacenters.rds")

cat("Wrote", nrow(final_df), "rows to outputs/malaysia_datacenters.csv\n")
```

# Reflection

This pilot produced a working dataset of datacenters in Malaysia built from
two openly-licensed sources with clean provenance. The multi-source
reconciliation approach turned out to be more analytically interesting than
pulling from a single directory would have been — the disagreement between
OSM and Wikidata on which facilities "count" is itself informative about
coverage gaps in open geospatial data.

For the global scale-up, the Overpass query can be modified by removing the
`area["ISO3166-1"="MY"]` filter or swapping in a different ISO code, though
a global query will likely need to be chunked by region to avoid Overpass
timeouts. The Wikidata query scales trivially by removing the `wdt:P17 wd:Q833`
constraint. The cloud provider JSON endpoints (AWS, Azure, GCP) are the
natural next layer to add for hyperscale coverage, since OSM tends to
under-represent hyperscale facilities that are unmarked on the ground.
