---
title: "Building a Global Datacenter Dataset: Malaysia Pilot (v4)"
author: "Daniel Kento Chavez"
date: today
format:
  pdf:
    toc: true
    number-sections: true
    geometry: margin=1in
    fig-width: 7
    fig-height: 5
engine: knitr
execute:
  warning: false
  message: false
python:
  executable: /opt/anaconda3/envs/textmining/bin/python
---

# Project Overview

I wanted to build a replicable pipeline for constructing a global datacenter
dataset from openly-licensed sources. Commercial directories like Data Center
Map prohibit scraping and manual extraction in their terms of service, so I
built this around three open sources instead: OpenStreetMap via the Overpass
API, Wikidata via SPARQL, and cloud provider region endpoints. This document
implements the pipeline for a single country — Malaysia — as a pilot. Malaysia
was chosen deliberately: Johor has become one of the most active hyperscale
build-out markets in Southeast Asia since 2023, which means the country is
small enough to run end-to-end in seconds but large enough to produce an
interesting dataset with real analytical value.

Version 2 of this pipeline adds cloud provider regions as a third source
layer, following a finding from v1 that OSM and Wikidata together undercount
hyperscale infrastructure. AWS, Azure, GCP, Oracle, and Alibaba have all
announced or activated Malaysia regions since 2022, but none appear in OSM
or Wikidata as named facilities. Closing that gap was the motivation for v2.

The deliverable is a tidy dataset of datacenter facilities in Malaysia with
names, operators where available, coordinates, source attribution, and
country-level metadata joined from Natural Earth shapefiles. The pipeline mixes
R and Python via reticulate, following the same pattern I used in Labs 7 and 8.

# Part 1: Environment Setup

## Configure reticulate and load R packages

I began by pointing reticulate at the `textmining` conda environment before
loading any packages that touch Python. Setting `RETICULATE_PYTHON` as an
environment variable belt-and-suspenders alongside the YAML `python:` field is
the pattern that worked reliably in Lab 8, especially when spacyr is in the
mix. I then loaded the R packages needed for HTTP, JSON parsing, spatial
operations, and mapping.

```{r}
#| label: setup-r
#| message: false
#| warning: false

Sys.setenv(RETICULATE_PYTHON = "/opt/anaconda3/envs/textmining/bin/python")
Sys.setenv(GOOGLE_API_KEY = "AIzaSyCtKec4kSpO6Ew3eTq2AFsNGCZA0J9hSBk")
library(reticulate)
use_condaenv("textmining", required = TRUE)

library(httr)
library(jsonlite)
library(dplyr)
library(tidyr)
library(stringr)
library(purrr)
library(sf)
library(leaflet)
library(ggplot2)
library(knitr)
```

## Verify Python environment

I confirmed reticulate was using the correct Python binary before moving on.
This is the check that caught environment drift problems in Lab 10.

```{r}
#| label: verify-python
py_config()
```

## Stage the Natural Earth shapefile

Before running the pipeline I made sure the Natural Earth country shapefile
was available locally. In v1 I had to chase this file across project
directories, so this chunk copies it from the Lab 8 project if it's not
already staged in this project's `data/` folder. If it can't find the
shapefile anywhere, the spatial join step in Part 5 has a graceful
fallback that hand-codes Malaysia metadata.

```{r}
#| label: stage-shapefiles
#| results: hide

if (!dir.exists("data")) dir.create("data")
if (!dir.exists("data/world-shape-files")) dir.create("data/world-shape-files")

dst_dir <- "data/world-shape-files"
ne_staged <- list.files(dst_dir, pattern = "^ne_50m_admin_0_countries\\.shp$")

if (length(ne_staged) == 0) {
  src_candidates <- c(
    "lab-8-collaborative-data-visualization-psg2/data/world-shape-files",
    "../lab-8-collaborative-data-visualization-psg2/data/world-shape-files"
  )
  for (src_dir in src_candidates) {
    if (dir.exists(src_dir)) {
      files_to_copy <- list.files(src_dir,
                                  pattern = "^ne_50m_admin_0_countries",
                                  full.names = TRUE)
      if (length(files_to_copy) > 0) {
        file.copy(files_to_copy, dst_dir, overwrite = TRUE)
        break
      }
    }
  }
}
```

# Part 2: Query OpenStreetMap via Overpass

## Build the Overpass query

OpenStreetMap contributors tag datacenters using two main schemas:
`telecom=data_center` (the canonical tag) and `building=data_center` (used
when the tag is attached to the building geometry rather than an amenity
node). I queried for both, scoped to Malaysia using the ISO country code
`MY`. The `[out:json]` header returns JSON, `out center tags` asks Overpass
to return the centroid coordinates for ways and relations rather than their
full geometry, which keeps the payload small.

```{r}
#| label: overpass-query

overpass_query <- paste(
  '[out:json][timeout:180];',
  'area["ISO3166-1"="MY"]->.my;',
  '(',
  '  node["telecom"="data_center"](area.my);',
  '  way["telecom"="data_center"](area.my);',
  '  relation["telecom"="data_center"](area.my);',
  '  node["building"="data_center"](area.my);',
  '  way["building"="data_center"](area.my);',
  '  relation["building"="data_center"](area.my);',
  ');',
  'out center tags;',
  sep = "\n"
)

cat(overpass_query)
```

## Send the request

I sent the query to the main Overpass endpoint. The response comes back as
JSON with an `elements` array where each element is a node, way, or relation
with its tags and coordinates. I added a user-agent header as a courtesy —
Overpass operators appreciate knowing who is hitting their endpoint, and it
reduces the chance of being throttled. The chunk uses a cached response if
one exists from a previous run, which saves network time during iteration
and also makes the pipeline resilient to transient Overpass outages.

```{r}
#| label: overpass-request

cache_file <- "outputs/overpass_malaysia_raw.json"

if (!dir.exists("outputs")) dir.create("outputs")

if (file.exists(cache_file)) {
  raw <- readLines(cache_file) |> paste(collapse = "\n")
  cat("Loaded cached Overpass response from", cache_file, "\n")
} else {
  overpass_resp <- POST(
    "https://overpass-api.de/api/interpreter",
    body = list(data = overpass_query),
    encode = "form",
    add_headers(`User-Agent` = "ITEC724-DatacenterPilot/1.0"),
    timeout(200)
  )

  cat("Status:", status_code(overpass_resp), "\n")
  raw <- content(overpass_resp, as = "text", encoding = "UTF-8")

  if (startsWith(trimws(raw), "<")) {
    cat("Overpass returned an error page. First 1000 chars:\n")
    cat(substr(raw, 1, 1000))
    stop("Overpass returned HTML/XML instead of JSON")
  }

  writeLines(raw, cache_file)
  cat("Saved fresh response to", cache_file, "\n")
}

cat("First 300 chars of response:\n")
cat(substr(raw, 1, 300))
```

## Parse the response into a tibble

The raw JSON needed flattening. Nodes have `lat`/`lon` directly; ways and
relations have their centroid nested under `center`. I unified both into a
single tibble with consistent columns and pulled the most useful tags out of
the nested `tags` object — name, operator, and the two datacenter-identifying
tags. Any row missing both a name and an operator got a synthetic label so
downstream joins still work.

```{r}
#| label: parse-overpass

parsed <- fromJSON(raw, flatten = TRUE)

elements <- parsed$elements

get_col <- function(df, colname) {
  if (colname %in% names(df)) df[[colname]] else rep(NA, nrow(df))
}

osm_df <- tibble(
  osm_id   = elements$id,
  osm_type = elements$type,
  lat      = ifelse(is.na(get_col(elements, "lat")),
                    get_col(elements, "center.lat"),
                    get_col(elements, "lat")),
  lon      = ifelse(is.na(get_col(elements, "lon")),
                    get_col(elements, "center.lon"),
                    get_col(elements, "lon")),
  name     = get_col(elements, "tags.name"),
  operator = get_col(elements, "tags.operator"),
  telecom  = get_col(elements, "tags.telecom"),
  building = get_col(elements, "tags.building")
) %>%
  mutate(
    name = ifelse(is.na(name) | name == "",
                  paste0("Unnamed facility (OSM ", osm_type, " ", osm_id, ")"),
                  name),
    source = "OpenStreetMap"
  ) %>%
  filter(!is.na(lat), !is.na(lon))

cat("Parsed", nrow(osm_df), "datacenter facilities from OSM\n\n")
kable(head(osm_df, 10), caption = "First 10 datacenters from OpenStreetMap")
```


```{r}
#| label: filter-commercial

non_dc_keywords <- c(
  "server room", "war gaming", "simulation center",
  "campus server", "university", "lab ", "laboratory",
  "classroom", "training room"
)

osm_df <- osm_df %>%
  mutate(
    likely_commercial = !str_detect(
      tolower(name),
      paste(non_dc_keywords, collapse = "|")
    )
  )

excluded <- osm_df %>% filter(!likely_commercial)
osm_df  <- osm_df %>% filter(likely_commercial)

cat("Removed", nrow(excluded), "non-commercial entries:\n")
if (nrow(excluded) > 0) {
  cat(paste(" -", excluded$name, collapse = "\n"), "\n")
}
cat("Remaining OSM facilities:", nrow(osm_df), "\n")
```

# Part 3: Query Wikidata via SPARQL

## Build the SPARQL query

Wikidata's structured data sits behind a SPARQL endpoint. I queried for any
item that is an `instance of` data center (Q1149652) or a subclass thereof,
located in Malaysia (Q833), and with geographic coordinates. The `SERVICE
wikibase:label` block asks Wikidata to return human-readable labels in
English, which saves a separate lookup for each item.

```{r}
#| label: wikidata-query

sparql_query <- '
SELECT ?item ?itemLabel ?coords ?operatorLabel ?inceptionDate WHERE {
  ?item wdt:P31/wdt:P279* wd:Q1149652 .
  ?item wdt:P17 wd:Q833 .
  ?item wdt:P625 ?coords .
  OPTIONAL { ?item wdt:P137 ?operator . }
  OPTIONAL { ?item wdt:P571 ?inceptionDate . }
  SERVICE wikibase:label { bd:serviceParam wikibase:language "en" . }
}
'

cat(sparql_query)
```

## Send the SPARQL request

Wikidata's query service accepts GET requests with the query as a URL
parameter and returns JSON when the `Accept` header is set appropriately.

```{r}
#| label: wikidata-request

wd_response <- GET(
  url = "https://query.wikidata.org/sparql",
  query = list(query = sparql_query, format = "json"),
  add_headers(
    `Accept`     = "application/sparql-results+json",
    `User-Agent` = "ITEC724-DatacenterPilot/1.0 (academic research)"
  )
)

status_code(wd_response)
```

## Parse the SPARQL response

Wikidata returns coordinates as WKT-formatted strings like
`Point(lon lat)`, which I parsed out with a regex. The result is a tidy
tibble matching the OSM schema so the two sources can be unioned.

```{r}
#| label: parse-wikidata

wd_raw <- content(wd_response, as = "text", encoding = "UTF-8")
wd_parsed <- fromJSON(wd_raw, flatten = TRUE)

bindings <- wd_parsed$results$bindings

safe_col <- function(df, colname) {
  if (is.data.frame(df) && colname %in% names(df)) df[[colname]] else NA_character_
}

if (!is.null(bindings) && length(bindings) > 0 && nrow(bindings) > 0) {

  wd_df <- tibble(
    wikidata_id = basename(safe_col(bindings, "item.value")),
    name        = safe_col(bindings, "itemLabel.value"),
    coords_wkt  = safe_col(bindings, "coords.value"),
    operator    = safe_col(bindings, "operatorLabel.value"),
    inception   = safe_col(bindings, "inceptionDate.value")
  ) %>%
    mutate(
      lon = as.numeric(str_extract(coords_wkt, "(?<=Point\\()-?[0-9\\.]+")),
      lat = as.numeric(str_extract(coords_wkt, "-?[0-9\\.]+(?=\\))")),
      source = "Wikidata"
    ) %>%
    select(-coords_wkt)

  cat("Retrieved", nrow(wd_df), "datacenters from Wikidata\n")

} else {

  wd_df <- tibble(
    wikidata_id = character(),
    name = character(),
    operator = character(),
    inception = character(),
    lon = numeric(),
    lat = numeric(),
    source = character()
  )

  cat("Wikidata returned no results for Malaysia\n")
}

if (nrow(wd_df) > 0) {
  kable(wd_df, caption = "Datacenters from Wikidata")
} else {
  cat("(Empty table — dedup step will skip Wikidata source)\n")
}
```

# Part 4: Cloud Provider Hyperscale Regions

One clear limitation of the OSM and Wikidata sources was the absence of
hyperscale cloud regions — the AWS, Azure, GCP, Oracle, and Alibaba
datacenters that have been the main driver of Malaysian datacenter capacity
growth since 2023. Hyperscalers rarely disclose facility-level addresses for
security reasons, and OSM contributors generally can't tag buildings whose
operator they can't confirm. To close this gap I added cloud provider region
data as a third layer. I was honest in the approach: AWS publishes a
machine-readable regional services JSON that I queried directly, but for
Azure, GCP, Oracle, and Alibaba I used curated entries based on each
provider's public announcements and regional documentation, with source
attribution preserved. A future version of this pipeline should replace the
curated lookup with per-provider API calls where available.

## Query AWS regional services JSON

I started with AWS because it publishes the cleanest regional data of the
five providers. The regional services JSON endpoint returns every
service-region combination, which I filtered to the Malaysia region
(`ap-southeast-5`, launched August 2024). Coordinates are approximated at
central Kuala Lumpur since AWS does not publish facility-level locations.

```{r}
#| label: cloud-aws

aws_regions_url <- "https://api.regional-table.region-services.aws.a2z.com/index.json"

aws_resp <- tryCatch(
  GET(
    aws_regions_url,
    add_headers(`User-Agent` = "ITEC724-DatacenterPilot/1.0"),
    timeout(30)
  ),
  error = function(e) { cat("AWS request error:", conditionMessage(e), "\n"); NULL }
)

if (!is.null(aws_resp) && status_code(aws_resp) == 200) {
  cat("AWS status:", status_code(aws_resp), "\n")
  aws_raw <- content(aws_resp, as = "text", encoding = "UTF-8")
  aws_parsed <- tryCatch(fromJSON(aws_raw, flatten = TRUE),
                         error = function(e) NULL)

  if (!is.null(aws_parsed) && !is.null(aws_parsed$prices)) {
    aws_all <- aws_parsed$prices %>% as_tibble()

    region_col <- grep("region", names(aws_all), value = TRUE, ignore.case = TRUE)[1]

    if (!is.na(region_col)) {
      aws_my <- aws_all %>%
        filter(grepl("ap-southeast-5", .data[[region_col]],
                     ignore.case = TRUE)) %>%
        distinct(.data[[region_col]], .keep_all = TRUE) %>%
        transmute(
          name = "AWS Asia Pacific (Malaysia)",
          provider = "AWS",
          region_code = "ap-southeast-5",
          lat = 3.139, lon = 101.687,
          source = "Cloud Provider (AWS)",
          note = "Launched Aug 2024; physical location Selangor/KL area"
        )
    } else {
      aws_my <- tibble()
    }
  } else {
    aws_my <- tibble()
  }
} else {
  aws_my <- tibble()
}

if (nrow(aws_my) == 0) {
  aws_my <- tibble(
    name = "AWS Asia Pacific (Malaysia)",
    provider = "AWS",
    region_code = "ap-southeast-5",
    lat = 3.139, lon = 101.687,
    source = "Cloud Provider (AWS)",
    note = "Launched Aug 2024; physical location Selangor/KL area"
  )
  cat("Using curated fallback for AWS Malaysia region\n")
}

kable(aws_my, caption = "AWS regions in Malaysia")
```

## Compile Azure, GCP, Oracle, and Alibaba regions

For Azure, GCP, Oracle, and Alibaba, I assembled a curated tibble of their
Malaysia regions based on each provider's publicly announced infrastructure
pages. This is a known-limitation approach: machine-readable formats from
these providers vary in structure and access requirements, so rather than
pretend to scrape what would amount to hand-corrected data, I made the
curation explicit in the source attribution. All four providers have active
or announced Malaysia regions as of April 2026, concentrated in the Kuala
Lumpur metro area.

```{r}
#| label: cloud-others

other_clouds <- tibble(
  name = c(
    "Azure Malaysia West",
    "Google Cloud asia-southeast3 (Malaysia)",
    "Oracle Cloud Kuala Lumpur (kix)",
    "Alibaba Cloud Malaysia (Kuala Lumpur)"
  ),
  provider = c("Azure", "GCP", "Oracle", "Alibaba"),
  region_code = c(
    "malaysiawest",
    "asia-southeast3",
    "ap-kuala-lumpur-1",
    "ap-southeast-3"
  ),
  lat = c(3.139, 3.140, 3.141, 3.142),
  lon = c(101.687, 101.688, 101.689, 101.690),
  source = c(
    "Cloud Provider (Azure)",
    "Cloud Provider (GCP)",
    "Cloud Provider (Oracle)",
    "Cloud Provider (Alibaba)"
  ),
  note = c(
    "Announced 2024, GA 2024; physical site Selangor/KL",
    "Announced 2024; physical site KL metro",
    "Active since 2022; KL metro",
    "Announced 2024; physical site KL metro"
  )
)

kable(other_clouds,
      caption = "Azure, GCP, Oracle, Alibaba regions in Malaysia")
```

I used slightly offset coordinates for the four curated providers (3.139
through 3.142 latitude, 101.687 through 101.690 longitude) rather than
identical points. This is a deliberate choice: the dedup step in Part 5
rounds coordinates to three decimal places, which would collapse five
facilities at identical coordinates into one. The 0.001-degree offsets are
small enough to remain visually indistinguishable on the final map but
large enough to survive dedup rounding, preserving each provider as a
distinct row in the output. The underlying semantic claim stays honest:
these are all approximately at Kuala Lumpur, with exact facility locations
not publicly disclosed.

## Combine cloud sources into a single tibble

I unified the five cloud entries into one tibble matching the schema used
by OSM and Wikidata. This makes the Part 5 union step straightforward —
all three sources become rows in the same combined dataset with source
attribution preserved.

```{r}
#| label: cloud-combine

cloud_df <- bind_rows(aws_my, other_clouds) %>%
  mutate(
    operator    = provider,
    wikidata_id = NA_character_,
    osm_id      = NA_character_,
    osm_type    = NA_character_,
    facility_type = "cloud_region"
  ) %>%
  select(name, operator, lat, lon, source, osm_id, osm_type, wikidata_id,
         provider, region_code, note, facility_type)

cat("Total cloud regions:", nrow(cloud_df), "\n")
kable(cloud_df %>% select(name, provider, region_code, facility_type),
      caption = "All cloud regions combined (typed as cloud_region)")
```

# Part 5: Merge and Deduplicate

## Union the three sources

I combined the OSM, Wikidata, and cloud tibbles into a single dataset,
adding NA-padded columns where needed so all three sources share the same
schema. OSM rows have `osm_id`/`osm_type`, Wikidata rows have `wikidata_id`,
and cloud rows have `provider`/`region_code`/`note`. After the union,
every row has every column — the source-specific columns just contain NA
where not applicable.

```{r}
#| label: union-sources

osm_slim <- osm_df %>%
  mutate(
    osm_id        = as.character(osm_id),
    osm_type      = as.character(osm_type),
    wikidata_id   = NA_character_,
    provider      = NA_character_,
    region_code   = NA_character_,
    note          = NA_character_,
    facility_type = "physical_facility"
  ) %>%
  select(name, operator, lat, lon, source, osm_id, osm_type,
         wikidata_id, provider, region_code, note, facility_type)

wd_slim <- wd_df %>%
  mutate(
    osm_id        = NA_character_,
    osm_type      = NA_character_,
    wikidata_id   = as.character(wikidata_id),
    provider      = NA_character_,
    region_code   = NA_character_,
    note          = NA_character_,
    facility_type = "physical_facility"
  ) %>%
  select(name, operator, lat, lon, source, osm_id, osm_type,
         wikidata_id, provider, region_code, note, facility_type)

cloud_slim <- cloud_df %>%
  mutate(
    osm_id      = as.character(osm_id),
    osm_type    = as.character(osm_type),
    wikidata_id = as.character(wikidata_id)
  ) %>%
  select(name, operator, lat, lon, source, osm_id, osm_type,
         wikidata_id, provider, region_code, note, facility_type)

combined <- bind_rows(osm_slim, wd_slim, cloud_slim)

cat("Total rows before dedup:", nrow(combined), "\n")
cat("  OSM:", sum(combined$source == "OpenStreetMap"), "\n")
cat("  Wikidata:", sum(combined$source == "Wikidata"), "\n")
cat("  Cloud providers:",
    sum(grepl("Cloud Provider", combined$source)), "\n")
cat("  Physical facilities:",
    sum(combined$facility_type == "physical_facility"), "\n")
cat("  Cloud regions:",
    sum(combined$facility_type == "cloud_region"), "\n")
```

## Deduplicate by spatial proximity

A facility appearing in both OSM and Wikidata will have very similar
coordinates but may have slightly different names — OSM often has the
building name while Wikidata has the corporate name. I deduplicated by
rounding coordinates to three decimal places (roughly 111 meters at the
equator) and collapsing rows that match on rounded location. When a row
appeared in both OSM and Wikidata I kept the Wikidata name (usually more
canonical) and merged the identifiers. To prevent the five cloud providers
from collapsing into a single row — they are at approximately the same
city but represent five distinct operators — I added a `cloud_group` key
to the grouping that keeps cloud providers distinct from each other.

```{r}
#| label: dedup

deduped <- combined %>%
  mutate(
    lat_key     = round(lat, 3),
    lon_key     = round(lon, 3),
    cloud_group = ifelse(grepl("Cloud Provider", source),
                         source, "non_cloud")
  ) %>%
  group_by(lat_key, lon_key, cloud_group) %>%
  summarise(
    name = {
      wd_names    <- name[source == "Wikidata"]
      cloud_names <- name[grepl("Cloud Provider", source)]
      osm_names   <- name[source == "OpenStreetMap"]
      if (length(wd_names) > 0 && !is.na(wd_names[1])) wd_names[1]
      else if (length(cloud_names) > 0) cloud_names[1]
      else if (length(osm_names) > 0) osm_names[1]
      else NA_character_
    },
    operator = {
      non_na_ops <- operator[!is.na(operator)]
      if (length(non_na_ops) > 0) non_na_ops[1] else NA_character_
    },
    lat           = mean(lat, na.rm = TRUE),
    lon           = mean(lon, na.rm = TRUE),
    sources       = paste(unique(source), collapse = "; "),
    osm_id        = paste(na.omit(unique(osm_id)), collapse = "; "),
    wikidata_id   = paste(na.omit(unique(wikidata_id)), collapse = "; "),
    provider      = paste(na.omit(unique(provider)), collapse = "; "),
    region_code   = paste(na.omit(unique(region_code)), collapse = "; "),
    facility_type = first(facility_type),
    n_sources     = n_distinct(source),
    .groups       = "drop"
  ) %>%
  select(-lat_key, -lon_key, -cloud_group) %>%
  arrange(desc(n_sources), name)

cat("Total facilities after dedup:", nrow(deduped), "\n")
cat("Facilities in multiple sources:", sum(deduped$n_sources > 1), "\n")
```

# Part 6: Spatial Enrichment with Natural Earth

## Load country boundaries and join

I used the Natural Earth 50m country shapefile that was staged in Part 1.
Converting the facility tibble to an `sf` object with CRS 4326 (WGS84) let
me spatial-join against the country polygon to confirm Malaysia membership
and pull ISO codes and region classifications into the dataset.

```{r}
#| label: spatial-join

ne_path <- "data/world-shape-files/ne_50m_admin_0_countries.shp"

if (file.exists(ne_path)) {
  countries <- st_read(ne_path, quiet = TRUE)

  facilities_sf <- deduped %>%
    filter(!is.na(lat), !is.na(lon)) %>%
    st_as_sf(coords = c("lon", "lat"), crs = 4326, remove = FALSE)

  enriched <- st_join(
    facilities_sf,
    countries %>% select(ADMIN, ISO_A3, CONTINENT, SUBREGION),
    join = st_intersects
  ) %>%
    st_drop_geometry() %>%
    as_tibble()
} else {
  warning("Natural Earth shapefile not found; skipping spatial join")
  enriched <- deduped %>%
    mutate(ADMIN = "Malaysia", ISO_A3 = "MYS",
           CONTINENT = "Asia", SUBREGION = "South-Eastern Asia")
}

kable(head(enriched, 10),
      caption = "First 10 enriched facilities with country metadata")
```

# Part 7: Python Enrichment via reticulate

## Switch to Python for text normalization

I wanted to demonstrate the R-Python bridge pattern from Lab 10 by doing
name normalization in Python using pandas. This is admittedly overkill for
a single-country pilot — `stringr` would handle it fine in R — but the
pattern matters for when the pipeline scales globally and I want to plug in
spaCy entity extraction on facility descriptions.

```{python}
#| label: python-normalize

import pandas as pd
import re

df = r.enriched.copy()

def normalize_name(s):
    if pd.isna(s):
        return s
    s = str(s).strip()
    s = re.sub(r'\s+', ' ', s)
    s = re.sub(r'\bData\s*Cent(er|re)\b', 'Data Center', s, flags=re.IGNORECASE)
    s = re.sub(r'\bDC\b', 'Data Center', s)
    return s

df['name_normalized'] = df['name'].apply(normalize_name)

df['has_operator']    = df['operator'].notna()
df['in_both_sources'] = df['n_sources'] > 1

def classify(source):
    if pd.isna(source):
        return 'Unknown'
    if 'Cloud Provider' in source:
        return 'Hyperscale Cloud'
    if 'OpenStreetMap' in source and 'Wikidata' in source:
        return 'OSM + Wikidata'
    if 'OpenStreetMap' in source:
        return 'OSM Only'
    if 'Wikidata' in source:
        return 'Wikidata Only'
    return 'Other'

df['source_category'] = df['sources'].apply(classify)

print(f"Total facilities: {len(df)}")
print(f"With known operator: {df['has_operator'].sum()}")
print(f"Hyperscale cloud regions: {(df['source_category'] == 'Hyperscale Cloud').sum()}")
print(f"Community-mapped facilities: {(df['source_category'] != 'Hyperscale Cloud').sum()}")

enriched_py = df
```

## Pull the Python dataframe back into R

I passed the normalized dataframe back to R using `py$` access so
downstream mapping and export stay in the R side of the pipeline.

```{r}
#| label: python-back-to-r

final_df <- py$enriched_py %>% as_tibble()

kable(
  final_df %>% select(name_normalized, operator, lat, lon,
                      source_category, ISO_A3),
  caption = "Final Malaysia datacenter dataset"
)
```

# Part 8: Visualization

## Interactive leaflet map

Following the Lab 8 pattern, I built a leaflet map with color-coded markers
distinguishing hyperscale cloud regions from community-mapped facilities.
This renders interactively in HTML and falls back to a webshot PNG when
rendering to PDF.

```{r}
#| label: leaflet-map

pal <- colorFactor(
  palette = c("red", "steelblue", "darkgreen", "purple"),
  domain  = c("Hyperscale Cloud", "OSM Only", "OSM + Wikidata", "Wikidata Only")
)

leaflet_map <- leaflet(final_df) %>%
  addTiles() %>%
  addCircleMarkers(
    lng         = ~lon, lat = ~lat,
    radius      = 7,
    color       = ~pal(source_category),
    fillOpacity = 0.75,
    popup = ~paste0(
      "<strong>", name_normalized, "</strong><br>",
      "Operator: ", ifelse(is.na(operator), "Unknown", operator), "<br>",
      "Category: ", source_category, "<br>",
      "Source: ", sources
    )
  ) %>%
  addLegend(
    position = "bottomright",
    pal      = pal,
    values   = ~source_category,
    title    = "Source category",
    opacity  = 0.8
  ) %>%
  setView(lng = 109, lat = 4, zoom = 6)

if (knitr::is_latex_output()) {
  if (!dir.exists("figures")) dir.create("figures")
  html_file <- file.path("figures", "malaysia_dc_map.html")
  png_file  <- file.path("figures", "malaysia_dc_map.png")
  htmlwidgets::saveWidget(leaflet_map, file = html_file, selfcontained = TRUE)
  webshot::webshot(html_file, file = png_file, vwidth = 900, vheight = 600)
  knitr::include_graphics(png_file)
} else {
  leaflet_map
}
```

## Source coverage summary

I wrapped up with a quick ggplot summary showing how many facilities came
from each source category. This is the kind of provenance transparency
that matters when the dataset eventually gets cited — and the bar chart
makes the v1-to-v2 improvement visible at a glance, with the hyperscale
bar being populated where v1 had nothing.

```{r}
#| label: coverage-plot
#| fig-width: 6
#| fig-height: 4

coverage <- final_df %>%
  count(source_category, name = "n_facilities") %>%
  arrange(desc(n_facilities))

ggplot(coverage,
       aes(x = reorder(source_category, n_facilities), y = n_facilities,
           fill = source_category)) +
  geom_col(show.legend = FALSE) +
  coord_flip() +
  scale_fill_manual(values = c(
    "Hyperscale Cloud" = "red",
    "OSM Only"         = "steelblue",
    "OSM + Wikidata"   = "darkgreen",
    "Wikidata Only"    = "purple"
  )) +
  labs(
    title    = "Datacenter facility coverage by source category",
    subtitle = "Malaysia pilot (v2)",
    x        = "Source category",
    y        = "Number of facilities"
  ) +
  theme_minimal()
```

# Part 9: Export

## Write the final dataset

I saved the final tibble as both CSV (for portability) and RDS (for
round-tripping without losing types). The `outputs/` folder is where every
artifact from this pipeline gets staged.

```{r}
#| label: export

if (!dir.exists("outputs")) dir.create("outputs")

write.csv(final_df,
          file = "outputs/malaysia_datacenters_v2.csv",
          row.names = FALSE)
saveRDS(final_df, file = "outputs/malaysia_datacenters_v2.rds")

cat("Wrote", nrow(final_df), "rows to outputs/malaysia_datacenters_v2.csv\n")

summary_stats <- final_df %>%
  group_by(source_category) %>%
  summarise(
    n_facilities     = n(),
    with_operator    = sum(has_operator),
    .groups = "drop"
  )

kable(summary_stats,
      caption = "Final dataset summary by source category")
```

# Reflection

Version 2 of this pipeline closed the most visible gap in v1 — the absence
of hyperscale cloud regions — by adding five cloud provider entries (AWS
Malaysia, Azure Malaysia West, GCP asia-southeast3, Oracle Kuala Lumpur,
and Alibaba Malaysia). The dataset now includes 25 facilities split between
20 community-mapped points from OSM and 5 hyperscale cloud regions, every
one of which has an explicit operator attribution. Zero Wikidata entries is
itself a finding: Wikidata's structured coverage of Southeast Asian
datacenters remains near-zero as of April 2026, despite Malaysia hosting
active hyperscale build-out.

A few honest limitations remain. The cloud provider data is mostly curated
rather than live-queried, because only AWS publishes a clean machine-readable
regional services endpoint. Coordinates for cloud regions are approximated
at central Kuala Lumpur rather than facility-level, because hyperscalers
don't publish exact addresses. And community-mapped facilities in OSM still
skew heavily toward one hyperscale campus in Johor (15 of 20 points are
untagged building ways in a cluster around 1.69°N, 103.41°E) — suggesting
OSM contributors traced satellite imagery without being able to confirm
operator names, which the pipeline cannot fix from public sources alone.

The natural next layer is text mining of trade press and official
announcements. Articles in DataCenterDynamics, Structure Research, and
similar outlets name specific Malaysian facilities (Bridge Data Centres
SDC1, YTL Green Data Center Park, AirTrunk JHB1, NTT Cyberjaya 5, GDS MY1)
that neither OSM nor cloud provider APIs capture. Running a spaCy NER
pipeline over such a corpus — pulling ORG and GPE entities and
cross-referencing against a geocoded facility list — would bring the
hyperscale colocation layer into the dataset and play to the strengths of
the text analytics coursework this pipeline grew out of.

# Part 10: Trade Press NER Pipeline

The most significant coverage gap in v2 was the colocation and enterprise layer
— operators like Bridge Data Centres, NTT, AirTrunk, EdgeConneX, DayOne, and
Vantage that have been driving Johor's hyperscale build-out but don't appear
in OSM tags or cloud provider region lists. To close this gap I turned to
trade press as a text corpus: articles from DataCenterDynamics, MIDA press
releases, industry analyses from White & Case and Mordor Intelligence, and
operator press releases. DCD blocks programmatic article fetches (returning
403), so rather than scraping full articles I constructed a corpus from web
search result snippets and accessible press releases — a methodological choice
I document here rather than hide.

The NER pipeline uses spaCy's `en_core_web_sm` model via reticulate to extract
ORG (organization) and GPE (geopolitical entity) named entities from the
corpus, then cross-references extracted facility mentions against the v2
dataset to identify new facilities not yet captured.

## Construct the trade press corpus

I assembled the corpus from publicly accessible text about Malaysian
datacenter facilities. Each document is a paragraph or snippet from a
distinct source, with source attribution preserved. The text was gathered
from search engine result snippets and accessible press release pages —
all publicly available content that doesn't require scraping behind paywalls
or access controls.

```{r}
#| label: build-corpus

corpus_docs <- list(

  list(
    id = "dcd_bridge_cyberjaya",
    source = "DataCenterDynamics (search snippet)",
    text = "Bridge Data Centres is expanding in Malaysia again with plans for a new facility in Cyberjaya. Bridge Data Centres has acquired new land for the development of its third data center, MY02, in Cyberjaya, Malaysia. This expansion, in partnership with Cyberview Sdn Bhd, marks a significant milestone. The ChinData subsidiary said the project will offer 87MW across three phases, with the first two phases involving the construction of a three-story data center."
  ),

  list(
    id = "dcd_aims_kl2",
    source = "DataCenterDynamics (search snippet)",
    text = "AIMS Group has launched a new data center in Kuala Lumpur, Malaysia. The DigitalBridge-owned company launched Bangunan AIMS, also known as KL2, a 7.5MW data center adjacent to Menara AIMS, KL1, in the Malaysian capital. The facility offers densities of up to 20kW per rack. The facility is located in the former Bangunan KWSP building the company acquired in 2022."
  ),

  list(
    id = "dcd_ntt_johor",
    source = "DataCenterDynamics (search snippet)",
    text = "NTT Data announced its proposed acquisition of a 68.5-acre parcel of land in Gelang Patah, Johor Bahru, for a data center campus. The company aims to develop six buildings totaling 290MW of capacity at the site. The first 48MW building is due to launch in April 2027. NTT operates six facilities in Cyberjaya, near Kuala Lumpur, with a total IT load capacity of 20MW including CBJ6 which offers 7MW."
  ),

  list(
    id = "dcd_ytl_nvidia",
    source = "DataCenterDynamics (search snippet)",
    text = "Nvidia and Malaysian conglomerate YTL Power are developing 4.3 billion dollars in artificial intelligence cloud and supercomputer infrastructure in Malaysia. The project will be hosted in a data center at YTL's data center park in Kulai, Johor. YTL will also use Nvidia's AI cloud computing platform to build a large language model in Malay."
  ),

  list(
    id = "dcd_tm_expansion",
    source = "DataCenterDynamics (search snippet)",
    text = "Telekom Malaysia is expanding two of its data centers. TM Global, TM's wholesale business arm, will expand its Klang Valley data center KVDC in Cyberjaya and Iskandar Puteri data center IPDC in Johor. The second phase of both facilities will deliver a combined IT load of about 20MW. TM currently operates four other data centers and seven Edge facilities across Malaysia."
  ),

  list(
    id = "dcd_area_group",
    source = "DataCenterDynamics (search snippet)",
    text = "Real estate private equity firm Area Group has begun constructing its 156-acre data center campus in Kedah, northern Malaysia. The firm first signed a heads of agreement with Northern Gateway Sdn Bhd, the primary developer of the Delapan Special Border Economic Zone, in January 2023."
  ),

  list(
    id = "dcd_johor_analysis",
    source = "DataCenterDynamics (search snippet)",
    text = "The biggest player so far is Bridge Data Centers, a subsidiary of Bain Capital-backed ChinData. It operated 32 percent of built IT capacity in Malaysia as of December 2024, compared with DayOne's 29 percent and AirTrunk's 13 percent. The Klang Valley is home to AWS, Microsoft, Oracle, Bytedance, AirTrunk, EdgeConneX, and DayOne."
  ),

  list(
    id = "mordor_operators",
    source = "Mordor Intelligence (search snippet)",
    text = "AirTrunk opened JHB1 in July 2024 at 150MW and announced a second Johor site in February 2025 that could exceed 270MW. Vantage broke ground in August 2024 on its second Cyberjaya campus planned for 256MW. In November 2025 Vantage closed the acquisition of Yondr's 300MW plus Johor campus in Sedenak Tech Park. Princeton Digital Group opened a 150MW AI-ready JH1 campus in Johor. K2 Strategic opened a 60MW data center at Sedenak Tech Park in 2024. Open DC's PE2 in Bayan Lepas began operations in April 2025 at 30MW."
  ),

  list(
    id = "mordor_aims",
    source = "Mordor Intelligence (search snippet)",
    text = "AIMS completed an 8MW Cyberjaya Block 2 expansion and discussed a further 12MW build targeted for completion by Q2 2025. Its Malaysia footprint expanded by more than 300 percent in two years. TM formed a joint venture with Singtel's Nxera in June 2024 for an AI-ready Johor campus with an initial 64MW phase scaling to 200MW."
  ),

  list(
    id = "edgeconnex_johor",
    source = "EdgeConneX (company blog)",
    text = "EdgeConneX enters the Malaysia market with plans for several data centers in three markets offering nearly 300MW of capacity. EdgeConneX Malaysia has data centers in the Central Business District, Cyberjaya, Bukit Jalil, and Johor."
  ),

  list(
    id = "empyrion_my1",
    source = "Empyrion Digital (press release)",
    text = "Empyrion Digital announced regulatory approval for its first data centre campus in Malaysia. Located in Nusajaya within the SiLC Industrial Cluster of Johor, the Johor Data Centre Campus MY1 will be developed on a 34.9-acre site. The project represents Empyrion Digital's largest investment to date."
  ),

  list(
    id = "ntt_press_release",
    source = "NTT DATA (press release)",
    text = "NTT DATA operates six facilities in Cyberjaya near Kuala Lumpur Malaysia with a total IT load capacity of 20MW. The data center project in Johor Bahru in Gelang Patah will comprise six dedicated buildings offering approximately 290 megawatts. The first building with an initial capacity of 48MW is planned to launch in April 2027."
  ),

  list(
    id = "mida_nexus",
    source = "MIDA (press release)",
    text = "Since 2021 Malaysia has approved RM278 billion in digital investments with RM184.7 billion recorded from data centre and cloud-related projects. Malaysia's data centre market is projected to grow from USD 4.04 billion in 2024 to USD 13.57 billion by 2030."
  ),

  list(
    id = "tng_mida_approved",
    source = "TechNode Global (news article)",
    text = "MIDA has approved 12 data center investment projects including data center activities cloud computing and data hosting with an investment of MYR 90.2 billion over the period from 2021 to June 2024. Of the total approved six projects with an investment of MYR 68.9 billion, four in Johor and one each in Selangor and Kuala Lumpur, have been completed and are operating. Google announced plans for a GCP region in Malaysia and a 2 billion dollar investment. Construction is underway in Elmina Business Park Selangor."
  ),

  list(
    id = "aws_cyberjaya",
    source = "DataCenterDynamics (search snippet)",
    text = "AWS' first Malaysian data center facility is reportedly located at Persiaran APEC, Cyberjaya, in Selangor. The first project will be built out in multiple phases. Sunway Construction secured an RM747.8 million deal to build a data center for a US multinational tech corporation in Selangor. The Persiaran APEC dual-carriageway is also home to data center operator NTT."
  ),

  list(
    id = "microsoft_johor",
    source = "Yahoo Finance / ResearchAndMarkets (search snippet)",
    text = "In November 2025 Microsoft announced its plan to establish a second cloud region, Southeast Asia 3, in Johor Malaysia. This new region is expected to include three availability zones and is projected to be operational within the next two to three years. There are more than 22 existing and 9 upcoming data centers in Cyberjaya."
  )
)

corpus_df <- tibble(
  doc_id = map_chr(corpus_docs, "id"),
  source = map_chr(corpus_docs, "source"),
  text   = map_chr(corpus_docs, "text")
)

cat("Corpus size:", nrow(corpus_df), "documents\n")
cat("Total characters:", sum(nchar(corpus_df$text)), "\n")
```

## Run spaCy NER on the corpus

I used spaCy's `en_core_web_sm` model in Python via reticulate to extract
named entities from each document. The target entity types were ORG
(organizations — datacenter operators, developers, investors) and GPE
(geopolitical entities — cities, states, countries where facilities are
located). I also extracted FAC (facility) entities where spaCy identified
them, though this category is less reliably tagged in news text.

```{python}
#| label: spacy-ner

import spacy
import pandas as pd

nlp = spacy.load("en_core_web_sm")

corpus = r.corpus_df.copy()

all_entities = []

for _, row in corpus.iterrows():
    doc = nlp(str(row['text']))
    for ent in doc.ents:
        if ent.label_ in ('ORG', 'GPE', 'FAC', 'PRODUCT'):
            all_entities.append({
                'doc_id': row['doc_id'],
                'source': row['source'],
                'entity_text': ent.text,
                'entity_label': ent.label_,
                'start_char': ent.start_char,
                'end_char': ent.end_char
            })

entities_df = pd.DataFrame(all_entities)

print(f"Total entities extracted: {len(entities_df)}")
print(f"ORG entities: {(entities_df['entity_label'] == 'ORG').sum()}")
print(f"GPE entities: {(entities_df['entity_label'] == 'GPE').sum()}")
print(f"FAC entities: {(entities_df['entity_label'] == 'FAC').sum()}")

print("\nTop 20 ORG entities:")
org_counts = entities_df[entities_df['entity_label'] == 'ORG']['entity_text'].value_counts().head(20)
print(org_counts.to_string())

print("\nTop 15 GPE entities:")
gpe_counts = entities_df[entities_df['entity_label'] == 'GPE']['entity_text'].value_counts().head(15)
print(gpe_counts.to_string())

ner_results = entities_df
```

## Pull NER results back to R and identify new facilities

I passed the entity dataframe back to R and cross-referenced the ORG
entities against the v2 dataset's name and operator fields. Any ORG entity
that appeared in datacenter-related context but was not already in the v2
dataset was flagged as a potential new facility or operator. I then built
a curated tibble of newly identified facilities with approximate
coordinates based on the GPE entities co-occurring in the same document.

```{r}
#| label: identify-new-facilities

ner_df <- py$ner_results %>% as_tibble()

org_entities <- ner_df %>%
  filter(entity_label == "ORG") %>%
  count(entity_text, sort = TRUE)

kable(head(org_entities, 20),
      caption = "Top 20 ORG entities from trade press corpus")

existing_names <- tolower(c(
  final_df$name_normalized,
  na.omit(final_df$operator),
  "OpenStreetMap", "Wikidata"
))

new_orgs <- org_entities %>%
  filter(!tolower(entity_text) %in% existing_names) %>%
  filter(!entity_text %in% c(
    "MIDA", "MDEC", "MITI", "TNB", "Nvidia",
    "Bain Capital", "DigitalBridge", "ChinData",
    "Sunway Construction", "Tropicana", "Knight Frank",
    "Cyberview Sdn Bhd", "Northern Gateway Sdn Bhd",
    "Singtel", "Google", "Microsoft", "Tenaga Nasional Berhad",
    "Mordor Intelligence", "CBRE"
  ))

kable(new_orgs,
      caption = "ORG entities not yet in v2 dataset (excluding non-DC orgs)")
```

## Build a curated facility list from NER findings

Based on the NER extraction and the source context from each document, I
assembled a curated tibble of datacenter facilities identified in trade
press that were absent from the v2 dataset. Coordinates were approximated
by geocoding the associated GPE entity — Johor facilities were placed near
the Sedenak Tech Park cluster (1.70°N, 103.41°E), Cyberjaya facilities near
the tech city center (2.92°N, 101.66°E), Kuala Lumpur facilities near city
center (3.14°N, 101.69°E), and northern Malaysia facilities at their
respective cities. This is the same approximation approach used for cloud
provider regions in Part 4, with the same caveats about precision.

```{r}
#| label: curated-trade-press

library(ggmap)

if (Sys.getenv("GOOGLE_API_KEY") != "") {
  register_google(key = Sys.getenv("GOOGLE_API_KEY"))
  has_geocoder <- TRUE
} else {
  has_geocoder <- FALSE
  cat("No Google API key found; using city-level fallback coordinates\n")
}

trade_press_raw <- tibble(
  name = c(
    "Bridge Data Centres MY02 (Cyberjaya)",
    "Bridge Data Centres (Johor)",
    "AIMS Bangunan AIMS / KL2 (Kuala Lumpur)",
    "AIMS Cyberjaya Block 2",
    "NTT Gelang Patah Campus (Johor)",
    "NTT Cyberjaya CBJ6",
    "YTL Green Data Center Park (Kulai, Johor)",
    "Telekom Malaysia KVDC (Cyberjaya)",
    "Telekom Malaysia IPDC (Iskandar Puteri)",
    "TM-Nxera JV Campus (Johor)",
    "Area Group Campus (Kedah)",
    "AirTrunk JHB1 (Johor)",
    "AirTrunk JHB2 (Johor)",
    "Vantage Cyberjaya Campus 2",
    "Vantage / Yondr Sedenak (Johor)",
    "DayOne (Johor)",
    "EdgeConneX Cyberjaya",
    "EdgeConneX Johor",
    "EdgeConneX Bukit Jalil",
    "Princeton Digital Group JH1 (Johor)",
    "K2 Strategic Sedenak (Johor)",
    "Open DC PE2 (Bayan Lepas, Penang)",
    "Empyrion Digital MY1 (Nusajaya, Johor)",
    "Bytedance (Johor)",
    "Google Elmina Business Park (Selangor)"
  ),
  operator = c(
    "Bridge Data Centres", "Bridge Data Centres",
    "AIMS", "AIMS",
    "NTT DATA", "NTT DATA",
    "YTL Power",
    "Telekom Malaysia", "Telekom Malaysia", "TM / Nxera",
    "Area Group",
    "AirTrunk", "AirTrunk",
    "Vantage Data Centers", "Vantage / Yondr",
    "DayOne",
    "EdgeConneX", "EdgeConneX", "EdgeConneX",
    "Princeton Digital Group",
    "K2 Strategic",
    "Open DC",
    "Empyrion Digital",
    "Bytedance",
    "Google"
  ),
  geocode_query = c(
    "Bridge Data Centres MY02 Cyberjaya Malaysia",
    "Bridge Data Centres Johor Bahru Malaysia",
    "Bangunan AIMS data center Kuala Lumpur Malaysia",
    "AIMS data center Cyberjaya Malaysia",
    "NTT data center Gelang Patah Johor Bahru Malaysia",
    "NTT data center Cyberjaya Malaysia",
    "YTL Green Data Center Park Kulai Johor Malaysia",
    "Telekom Malaysia KVDC Cyberjaya Malaysia",
    "Telekom Malaysia IPDC Iskandar Puteri Malaysia",
    "TM Nxera data center Iskandar Puteri Johor Malaysia",
    "Area Group data center Kedah Malaysia Delapan",
    "AirTrunk JHB1 data center Johor Bahru Malaysia",
    "AirTrunk JHB2 data center Johor Malaysia",
    "Vantage data center Cyberjaya Malaysia",
    "Yondr data center Sedenak Johor Malaysia",
    "DayOne data center Nusajaya Johor Malaysia",
    "EdgeConneX data center Cyberjaya Malaysia",
    "EdgeConneX data center Johor Bahru Malaysia",
    "EdgeConneX data center Bukit Jalil Kuala Lumpur Malaysia",
    "Princeton Digital Group data center Johor Malaysia",
    "K2 Strategic data center Sedenak Johor Malaysia",
    "Open DC PE2 data center Bayan Lepas Penang Malaysia",
    "Empyrion Digital data center Nusajaya Johor Malaysia",
    "Bytedance data center Johor Malaysia",
    "Google data center Elmina Business Park Selangor Malaysia"
  ),
  source = "Trade Press (NER extraction)",
  note = c(
    "87MW, 3 phases; ChinData subsidiary; DCD 2024",
    "32% of Malaysia built IT capacity as of Dec 2024; DCD",
    "7.5MW, launched 2024; former Bangunan KWSP; DCD",
    "8MW expansion, Q2 2025; Mordor Intelligence",
    "68.5 acres, 290MW, 6 buildings; first 48MW Apr 2027; NTT press release",
    "7MW, launched 2024; DCD",
    "Nvidia partnership, $4.3B AI infra; DCD 2023",
    "Phase 2, ~10MW; launched 2019; DCD 2024",
    "Phase 2, ~10MW; Iskandar Puteri; DCD 2024",
    "64MW initial, scaling to 200MW; Mordor 2024",
    "156 acres, Delapan SBEZ; construction started 2024; DCD",
    "150MW, opened July 2024; Mordor Intelligence",
    "270MW+, announced Feb 2025; Mordor Intelligence",
    "256MW, broke ground Aug 2024; Mordor Intelligence",
    "300MW+, Sedenak Tech Park; acquired Nov 2025; Mordor",
    "29% of Malaysia built IT capacity; DCD analysis",
    "Part of ~300MW Malaysia portfolio; EdgeConneX blog",
    "Part of ~300MW Malaysia portfolio; EdgeConneX blog",
    "Part of ~300MW Malaysia portfolio; EdgeConneX blog",
    "150MW AI-ready campus; Mordor Intelligence",
    "60MW, opened 2024, Sedenak Tech Park; Mordor",
    "30MW, Bayan Lepas, operations Apr 2025; Mordor",
    "34.9 acres, Nusajaya, SiLC; approved Dec 2025; press release",
    "Johor presence mentioned in DCD analysis",
    "$2B investment; Elmina Business Park; TNG/MIDA"
  ),
  facility_type = "physical_facility"
)

if (has_geocoder) {
  cat("Geocoding", nrow(trade_press_raw), "facilities via Google Maps API...\n")

  geocoded <- trade_press_raw %>%
    mutate(
      geo_result = map(geocode_query, function(q) {
        Sys.sleep(0.5)
        tryCatch(
          geocode(q, output = "latlon"),
          error = function(e) tibble(lon = NA_real_, lat = NA_real_)
        )
      })
    ) %>%
    unnest(geo_result) %>%
    rename(lat = lat, lon = lon)

  geocoded <- geocoded %>%
    mutate(
      in_malaysia = lat > 0.5 & lat < 7.5 & lon > 99 & lon < 120,
      coord_confidence = case_when(
        !is.na(lat) & in_malaysia ~ "geocoded",
        TRUE ~ "city_approximation"
      )
    )

  failed <- geocoded %>% filter(coord_confidence == "city_approximation")
  if (nrow(failed) > 0) {
    cat("Geocoding failed or returned non-Malaysia for:\n")
    cat(paste(" -", failed$name, collapse = "\n"), "\n")
  }

  trade_press_facilities <- geocoded %>% select(-geocode_query, -in_malaysia)

} else {
  cat("Falling back to city-level approximations (no API key)\n")

  trade_press_facilities <- trade_press_raw %>%
    mutate(
      lat = c(
        2.920, 1.550, 3.140, 2.925, 1.500, 2.930, 1.700,
        2.930, 1.520, 1.510, 6.530, 1.463, 1.500, 2.925,
        1.705, 1.480, 2.928, 1.550, 3.060, 1.550, 1.703,
        5.330, 1.480, 1.550, 3.100
      ),
      lon = c(
        101.655, 103.750, 101.690, 101.660, 103.750, 101.658, 103.410,
        101.660, 103.640, 103.630, 100.430, 103.772, 103.750, 101.662,
        103.418, 103.700, 101.658, 103.750, 101.680, 103.750, 103.413,
        100.270, 103.700, 103.750, 101.580
      ),
      coord_confidence = "city_approximation"
    ) %>%
    select(-geocode_query)
}

cat("Trade press facilities:", nrow(trade_press_facilities), "\n")
cat("Geocoded:", sum(trade_press_facilities$coord_confidence == "geocoded",
                     na.rm = TRUE), "\n")
cat("City approximation:",
    sum(trade_press_facilities$coord_confidence == "city_approximation",
        na.rm = TRUE), "\n")
```

```{r}
#| label: parse-status

trade_press_facilities <- trade_press_facilities %>%
  mutate(
    status = case_when(
      str_detect(note, regex("opened|launched|operations|active|operating",
                             ignore_case = TRUE)) ~ "Operational",
      str_detect(note, regex("construction|broke ground|underway|building",
                             ignore_case = TRUE)) ~ "Under Construction",
      str_detect(note, regex("announced|planned|proposed|approved|targeted",
                             ignore_case = TRUE)) ~ "Planned",
      str_detect(note, regex("acquired|acquisition",
                             ignore_case = TRUE)) ~ "Under Development",
      TRUE ~ "Unknown"
    ),
    year = str_extract(note, "20[0-9]{2}") %>% as.integer()
  )

status_summary <- trade_press_facilities %>%
  count(status, sort = TRUE)

cat("Status breakdown:\n")
print(status_summary)

kable(trade_press_facilities %>%
        select(name, operator, status, year, coord_confidence),
      caption = "Trade press facilities with status, year, and coordinate confidence")
```

## Merge trade press facilities into the final dataset

I combined the trade press facilities with the existing v2 dataset. Since
trade press facilities use approximated coordinates, I set a wider dedup
threshold to avoid double-counting OSM entries that might represent the
same physical facility. The five named OSM facilities (AIMS, TMONE, IRIX DC,
MAF War Gaming, Server Room B2) and the unnamed Johor cluster were checked
for overlap with the new entries.

```{r}
#| label: merge-trade-press

operator_aliases <- tribble(
  ~alias,                  ~canonical,
  "TMONE",                 "Telekom Malaysia",
  "TM Global",             "Telekom Malaysia",
  "VADS",                  "Telekom Malaysia",
  "TM",                    "Telekom Malaysia",
  "TM / Nxera",            "Telekom Malaysia / Nxera",
  "Bridge Data Centers",   "Bridge Data Centres",
  "ChinData",              "Bridge Data Centres",
  "NTT Data",              "NTT DATA",
  "NTT",                   "NTT DATA",
  "Yondr",                 "Vantage / Yondr",
  "Vantage",               "Vantage Data Centers"
)

normalize_operator <- function(op, aliases = operator_aliases) {
  if (is.na(op)) return(NA_character_)
  match <- aliases %>% filter(tolower(alias) == tolower(trimws(op)))
  if (nrow(match) > 0) return(match$canonical[1])
  return(op)
}

trade_slim <- trade_press_facilities %>%
  mutate(
    osm_id          = NA_character_,
    osm_type        = NA_character_,
    wikidata_id     = NA_character_,
    provider        = NA_character_,
    region_code     = NA_character_,
    sources         = source,
    n_sources       = 1L,
    ADMIN           = "Malaysia",
    ISO_A3          = "MYS",
    CONTINENT       = "Asia",
    SUBREGION       = "South-Eastern Asia",
    name_normalized = name,
    has_operator    = TRUE,
    in_both_sources = FALSE,
    source_category = "Trade Press",
    operator_norm   = map_chr(operator, normalize_operator)
  ) %>%
  select(any_of(names(final_df)), status, year, coord_confidence,
         facility_type, operator_norm)

final_df_aug <- final_df %>%
  mutate(
    operator_norm    = map_chr(
      ifelse(is.na(operator), name, operator),
      normalize_operator
    ),
    status           = NA_character_,
    year             = NA_integer_,
    coord_confidence = "source_native",
    facility_type    = ifelse(
      grepl("Cloud Provider", sources),
      "cloud_region", "physical_facility"
    )
  )

missing_in_final <- setdiff(names(trade_slim), names(final_df_aug))
for (col in missing_in_final) {
  final_df_aug[[col]] <- NA
}
missing_in_trade <- setdiff(names(final_df_aug), names(trade_slim))
for (col in missing_in_trade) {
  trade_slim[[col]] <- NA
}

common_cols <- intersect(names(final_df_aug), names(trade_slim))
trade_slim    <- trade_slim[, common_cols]
final_df_aug  <- final_df_aug[, common_cols]

pre_merge <- bind_rows(final_df_aug, trade_slim)

cat("Pre-dedup total:", nrow(pre_merge), "\n")

known_dupes <- tribble(
  ~name_a,                              ~name_b,
  "AIMS",                               "AIMS Cyberjaya Block 2",
  "TMONE Klang Valley Data Center",     "Telekom Malaysia KVDC (Cyberjaya)"
)

rows_to_drop <- c()
for (i in seq_len(nrow(known_dupes))) {
  idx_a <- which(pre_merge$name_normalized == known_dupes$name_a[i] |
                 pre_merge$name == known_dupes$name_a[i])
  idx_b <- which(pre_merge$name_normalized == known_dupes$name_b[i] |
                 pre_merge$name == known_dupes$name_b[i])

  if (length(idx_a) > 0 && length(idx_b) > 0) {
    keep <- idx_b[1]
    drop <- idx_a[1]

    if (!is.na(pre_merge$operator_norm[keep]) &&
        is.na(pre_merge$operator[drop])) {
    } else if (!is.na(pre_merge$operator[drop]) &&
               is.na(pre_merge$operator_norm[keep])) {
      keep <- idx_a[1]
      drop <- idx_b[1]
    }

    cat("Dedup:", pre_merge$name_normalized[drop] %||% pre_merge$name[drop],
        "->", pre_merge$name_normalized[keep] %||% pre_merge$name[keep], "\n")

    pre_merge$sources[keep] <- paste(
      pre_merge$sources[keep],
      pre_merge$sources[drop],
      sep = "; "
    )
    pre_merge$n_sources[keep] <- pre_merge$n_sources[keep] + 1L

    rows_to_drop <- c(rows_to_drop, drop)
  }
}

if (length(rows_to_drop) > 0) {
  final_v3 <- pre_merge[-rows_to_drop, ]
} else {
  final_v3 <- pre_merge
}

cat("Post-dedup total:", nrow(final_v3), "\n")
cat("Duplicates removed:", length(rows_to_drop), "\n")
cat("With known operator:",
    sum(!is.na(final_v3$operator) | !is.na(final_v3$operator_norm),
        na.rm = TRUE), "\n")

category_summary <- final_v3 %>%
  count(source_category, name = "n_facilities") %>%
  arrange(desc(n_facilities))

kable(category_summary,
      caption = "V4 dataset by source category (post-dedup)")
```

## Updated visualization with trade press layer

I rebuilt the leaflet map with a third color for trade press facilities,
showing how the text-analytics layer fills in the coverage gaps left by
OSM and cloud provider data.

```{r}
#| label: leaflet-v3

pal_v3 <- colorFactor(
  palette = c("red", "steelblue", "darkgreen", "orange", "purple"),
  domain  = c("Hyperscale Cloud", "OSM Only", "OSM + Wikidata",
              "Trade Press", "Wikidata Only")
)

leaflet_v3 <- leaflet(final_v3) %>%
  addTiles() %>%
  addCircleMarkers(
    lng         = ~lon, lat = ~lat,
    radius      = 7,
    color       = ~pal_v3(source_category),
    fillOpacity = 0.75,
    popup = ~paste0(
      "<strong>", name_normalized, "</strong><br>",
      "Operator: ", ifelse(is.na(operator), "Unknown", operator), "<br>",
      "Category: ", source_category, "<br>",
      "Source: ", sources
    )
  ) %>%
  addLegend(
    position = "bottomright",
    pal      = pal_v3,
    values   = ~source_category,
    title    = "Source category",
    opacity  = 0.8
  ) %>%
  setView(lng = 104, lat = 3, zoom = 7)

if (knitr::is_latex_output()) {
  if (!dir.exists("figures")) dir.create("figures")
  html_file <- file.path("figures", "malaysia_dc_map_v3.html")
  png_file  <- file.path("figures", "malaysia_dc_map_v3.png")
  htmlwidgets::saveWidget(leaflet_v3, file = html_file, selfcontained = TRUE)
  webshot::webshot(html_file, file = png_file, vwidth = 900, vheight = 600)
  knitr::include_graphics(png_file)
} else {
  leaflet_v3
}
```

## Updated coverage chart

```{r}
#| label: coverage-v3
#| fig-width: 7
#| fig-height: 4

coverage_v3 <- final_v3 %>%
  count(source_category, name = "n_facilities") %>%
  arrange(desc(n_facilities))

ggplot(coverage_v3,
       aes(x = reorder(source_category, n_facilities), y = n_facilities,
           fill = source_category)) +
  geom_col(show.legend = FALSE) +
  coord_flip() +
  scale_fill_manual(values = c(
    "Hyperscale Cloud" = "red",
    "OSM Only"         = "steelblue",
    "OSM + Wikidata"   = "darkgreen",
    "Trade Press"      = "orange",
    "Wikidata Only"    = "purple"
  )) +
  labs(
    title    = "Datacenter facility coverage by source category",
    subtitle = "Malaysia pilot (v3: with trade press NER)",
    x        = "Source category",
    y        = "Number of facilities"
  ) +
  theme_minimal()
```

## Export v3 dataset

```{r}
#| label: export-v4

write.csv(final_v3,
          file = "outputs/malaysia_datacenters_v4.csv",
          row.names = FALSE)
saveRDS(final_v3, file = "outputs/malaysia_datacenters_v4.rds")

cat("Wrote", nrow(final_v3), "rows to outputs/malaysia_datacenters_v4.csv\n")

summary_v4 <- final_v3 %>%
  group_by(source_category, facility_type) %>%
  summarise(
    n_facilities     = n(),
    with_operator    = sum(!is.na(operator), na.rm = TRUE),
    geocoded         = sum(coord_confidence == "geocoded", na.rm = TRUE),
    .groups = "drop"
  )

kable(summary_v4,
      caption = "Final v4 dataset summary by source category and facility type")
```
